Combinatorics of the geometry of Wilson loop diagrams I: equivalence classes via matroids and polytopes

### 2019-06-17 GWLDCI.tex
very minor edits (as discussed) to 1, incorporated Susama's changes to 2 as far as the end of section 4

A {\em matroid} $M = (E,\cB)$ consists of a finite ground set $E$ and a non-empty family $\cB \subseteq \cP(E)$ whose elements satisfy the {\em basis exchange property}: for any distinct $B_1,B_2 \in \cB$ and any $a \in B_1 \setminus B_2$, there exists some $b \in B_2 \setminus B_1$ such that $(B_1 \setminus \{a\})\cup \{b\} \in \cB$ as well. The elements of $\cB$ are called the {\em bases} of the matroid. Note that the basis exchange property immediately implies that all bases have the same size.

@@

\begin{lem}\label{lem facts about WLD matroids}
Let $W = (\cP,n)$ be an admissible Wilson loop diagram. Then:
\begin{enumerate}
\item The rank of a set $V \subseteq [n]$ is bounded above by $\min\{|V|,|\Prop(V)|\}$, with $\rk(V) = |V|$ if and only if $V$ is an independent set.
%\item Let $v \in [n]$ and $q \in \Prop(v)$. Then for any $V \subset [n]$ such that $q \not \in \Prop(V)$, we have $\rk(V \cup \{v\}) = \rk(V) +1$, i.e. adding a vertex that supports a new propagator to a set increases the rank of the set.
\item If $C \subseteq [n]$ is a cycle in $M(W)$, then $\rk(C) = |Prop(C)|$.
\item If $[n]$ can be partitioned into at least two non-empty sets, each of which support disjoint sets of propagators and in such a way that we obtain a partition of the propagator set, i.e. if we have \bas [n] = \sqcup_i V(P_i) \quad \textrm{ such that } \quad \sqcup P_i = \cP;  \text{ with }V(P_i) \cap V(P_j) = \emptyset \text{ and } P_i \cap P_j = \emptyset \quad \forall i \neq j\;, \eas then the matroid $M(W)$ is separable, i.e. \bas M(W) = \bigoplus_i M(P_i, V(P_i)) \;.\eas 
\item $F(P)$ is a flat of $M( W)$, which we call the \emph{propagator flat} of $P$.
\item If $F$ is a cyclic flat of $M(W)$, then it is a propagator flat.
\end{enumerate}
\end{lem}
\begin{proof}
The first part of (1) is \cite[Equation (9)]{wilsonloop} and surrounding discussion, and the second part is standard matroid theory. (2) is \cite[Lemma 3.27]{wilsonloop}. (3) is a direct consequence of \cite[Lemma 3.20]{wilsonloop} and the fact that $F(P_1)^c = V(P_1^c)$

@@

\begin{thm}\label{same matroid iff equiv}
Let $W= (\cP, [n])$ and $W'= (\cP', n)$ be two Wilson loop diagrams. They define the same matroid if and only if $W \sim W'$.
\end{thm}

@@

\begin{lem}\label{lem good quads}
Let $D_1$ and $D_2$ be two dissections of a convex $n$-gon defined by the vertices $x_1, \ldots, x_n$.  \hlfix{Suppose that $V_{>3}(D_1) = V_{>3}(D_2)$}{we probably use this implicitly somewhere in the proof but I can't see it; can we make it explicit?}, but that the $D_1$ and $D_2$ do not have the same set of non-triangle faces.  Then there are four vertices $x_i$, $x_j$, $x_k$, $x_l$ satisfying the following conditions:\begin{enumerate} \item In either $D_1$ or $D_2$ all four vertices are on the same non-triangle face; \item In the other dissection, there is a diagonal of the dissection such that exactly one of $\{x_i, x_j, x_k, x_l\}$ lies on one side of the diagonal (so the others are either end points of this diagonal or strictly on the other side). \end{enumerate}
\end{lem}

@@

  As observed in the proof of Proposition~\ref{prop easy way}, the support of any nonzero vector parallel to an edge is on the indices corresponding to the four vertices bounding the quadrilateral of the dissection associated with the edge.  Thus if $V_{>3}(\tau(W)) \neq V_{>3}(\tau(X))$ (notation as defined before Lemma~\ref{lem good quads}) then the span of the vectors parallel to their edges has different support and so they cannot be parallel.  So we can assume that $V_{>3}(\tau(W)) = V_{>3}(\tau(X))$.

@@

    Now, $\tau(W)$ and $\tau(X)$ satisfy the hypotheses of Lemma~\ref{lem good quads}, so we can assume that $f$ and $d$ have the property that we can choose a side of $d$ to be the positive side in such a way that exactly one vertex of $f$ is strictly on the positive side.  Then $\omega^+_d$ and the direction vector of $d$ have common support exactly on that one vertex and so their dot product is nonzero.  Since the direction vector of $d$ is contained in the direction of $\tau(X)$ but is not orthogonal to $\omega^+_d$, while $\omega^+_d$ is normal to the face of $\tau(W)$, it follows that the faces associated to $\tau(W)$ and $\tau(X)$ are not parallel, as desired.
\end{proof}

>>>>

A {\em matroid} $M = (E,\cB)$ consists of a finite ground set $E$ and a non-empty family $\cB$ of subsets of $E$, such that elements of $\cB$ satisfy the {\em basis exchange property}: for any distinct $B_1,B_2 \in \cB$ and any $a \in B_1 \setminus B_2$, there exists some $b \in B_2 \setminus B_1$ such that $(B_1 \setminus \{a\})\cup \{b\} \in \cB$ as well. The elements of $\cB$ are called the {\em bases} of the matroid. Note that the basis exchange property immediately implies that all bases have the same size.

@@

\begin{lem}\label{lem facts about WLD matroids}
Let $W = (\cP,[n])$ be an admissible Wilson loop diagram. Then:
\begin{enumerate}
\item The rank of a set $V \subseteq [n]$ is bounded above by $\min\{|V|,|\Prop(V)|\}$, with $\rk(V) = |V|$ if and only if $V$ is an independent set.
%\item Let $v \in [n]$ and $q \in \Prop(v)$. Then for any $V \subset [n]$ such that $q \not \in \Prop(V)$, we have $\rk(V \cup \{v\}) = \rk(V) +1$, i.e. adding a vertex that supports a new propagator to a set increases the rank of the set.
\item If $C \subseteq [n]$ is a cycle in $M(W)$, then $\rk(C) = |Prop(C)|$.
\item If $[n]$ can be partitioned into at least two non-empty sets, each of which support disjoint sets of propagators and in such a way that we obtain a partition of the propagator set, i.e. if we have \bas [n] = \sqcup_i V(P_i) \quad \textrm{ such that } \quad \sqcup P_i = \cP;  \text{ with }V(P_i) \cap V(P_j) = \emptyset \text{ and } P_i \cap P_j = \emptyset \quad \forall i \neq j\;, \eas then the matroid $M(W)$ is separable, i.e. \bas M(W) = \bigoplus_i M(P_i, V(P_i)) \;.\eas 
\item $F(P)$ is a flat of $M( W)$, which we call the \emph{propagator flat} of $P$.
\item If $F$ is a cyclic flat of $M(W)$, then it is a propagator flat.
\end{enumerate}
\end{lem}
\begin{proof}
The first part of (1) is \cite[Equation (9)]{wilsonloop} and surrounding discussion, and the second part is standard matroid theory. (2) is \cite[Lemma 3.27]{wilsonloop}. (3) is a direct consequence of \cite[Lemma 3.20]{wilsonloop} and the fact that $F(P_1)^c = V(P_1^c)$

@@

\begin{thm}\label{same matroid iff equiv}
Let $W= (\cP, [n])$ and $W'= (\cP', [n])$ be two Wilson loop diagrams. They define the same matroid if and only if $W \sim W'$.
\end{thm}

@@

\begin{lem}\label{lem good quads}
Let $D_1$ and $D_2$ be two dissections of a convex $n$-gon defined by the vertices $x_1, \ldots, x_n$.  Suppose that $V_{>3}(D_1) = V_{>3}(D_2)$, but that the $D_1$ and $D_2$ do not have the same set of non-triangle faces.  Then there are four vertices $x_i$, $x_j$, $x_k$, $x_l$ satisfying the following conditions:\begin{enumerate} \item In either $D_1$ or $D_2$ all four vertices are on the same non-triangle face; \item In the other dissection, there is a diagonal of the dissection such that exactly one of $\{x_i, x_j, x_k, x_l\}$ lies on one side of the diagonal (so the others are either end points of this diagonal or strictly on the other side). \end{enumerate}
\end{lem}

@@

  As observed in the proof of Proposition~\ref{prop easy way}, the support of any non-zero vector parallel to an edge is on the indices corresponding to the four vertices bounding the quadrilateral of the dissection associated with the edge.  Thus if $V_{>3}(\tau(W)) \neq V_{>3}(\tau(X))$ (notation as defined before Lemma~\ref{lem good quads}) then the span of the vectors parallel to their edges has different support and so they cannot be parallel.  So we can assume that $V_{>3}(\tau(W)) = V_{>3}(\tau(X))$.

@@

    Now, $\tau(W)$ and $\tau(X)$ satisfy the hypotheses of Lemma~\ref{lem good quads}, so we can assume that $f$ and $d$ have the property that we can choose a side of $d$ to be the positive side in such a way that exactly one vertex of $f$ is strictly on the positive side.  Then $\omega^+_d$ and the direction vector of $d$ have common support exactly on that one vertex and so their dot product is non-zero.  Since the direction vector of $d$ is contained in the direction of $\tau(X)$ but is not orthogonal to $\omega^+_d$, while $\omega^+_d$ is normal to the face of $\tau(W)$, it follows that the faces associated to $\tau(W)$ and $\tau(X)$ are not parallel, as desired.
\end{proof}
